---
title: "Centre for demography R for beginners"
author: 'Authors: Sébastien Lavoie, Elham Sirag'
date: 'Version: Sept 8th, 2022'
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```

## R syntax

### Review the basics

In this quiz, we will review the basic elements of R syntax that were covered at the beginning of our session. If you are unsure, 
you can always go review the lesson or use google. Don't stress, you are allowed as many tries as you want. 

```{r basics_quiz, echo=FALSE}
quiz(
  question_numeric(
    "In R, what would be the result of (6a + 42) / (3 * 5) if a = 3?",
    answer(4, correct = TRUE),
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that R follows standard mathematical oder of operations."
  ), 
  
  question_radio(
    "What symbol(s) is used to mark a line as containing comments in R?",
    answer("*"),
    answer("/*"),
    answer("#", correct = TRUE),
    answer("//"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  
  question_checkbox(
    "Select all that would be interpreted as valid strings (combination of characters) when typed in R as shown below. ",
    answer("'hello'", correct = TRUE),
    answer("hello"),
    answer("'déjà vu'", correct = TRUE),
    answer("three dogs"),
    answer("'32'", correct = TRUE),
    answer("3.1415"),
    random_answer_order = TRUE,
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that R needs string sto be clearly identified."
  ),
  
  question_checkbox(
    "Which two of the following are valid R vectors?",
    answer("c(1965, 12, 27)", correct = TRUE),
    answer("('1976', '08' '14')"),
    answer("c(1958, '06', 23)"),
    answer("c('1971', '02', '12')", correct = TRUE),
    answer("c(1986, '8', '6')"),
    random_answer_order = TRUE,
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that vectors must contain elements of the same type separated by commas AND inside the combine operator c()."
  )
  
)

```


### Vector concepts

For this exercise, you will be writing the code to review vector concepts. In the code box below:

1. Create a `phone_numbers` vector containing three different, 10-digit phone numbers as numeric values. <br>
2. Use `print()` to display the vector content. <br>
3. Using the square brackets `[]` inside the `print()` statement, display only the third element of the vector. <br>
4. **TIP :** you can use vectors `c()` inside square brackets `[]` to access more than one element at a time. Try it for yourself! Using the following syntax will display both the first and third element of your vector: `print(phone_numbers[c(1, 3)])`.

**To see the outputs of your code, click the `>Run Code` button.** 

Note: if you are stuck, click on the hint button.

```{r vectors_ex, exercise=TRUE, exercise.lines = 5}

```

```{r vectors_ex-hint}
# 1
phone_numbers <- c(1234567890, 0987654321, 5647382910)

# 2
print(phone_numbers)

# 3
print(phone_numbers[3])

# 4
print(phone_numbers[c(1, 3)])


```



## If-else statements and operators

Now that we've reviewed some of the basic syntax, let's move on to `if-else` statements with comparison and logical operators.

Comparison operators, as their name suggest, are used to evaluate one element against another. The comparison is evaluated like text
from left to right, returning either `TRUE` or `FALSE`. Comparison operators include `<`, `>`, `<=`, `>=`, `==` and `!=` (less than, greater than, less than or equal, greater than or equal, equal and not equal). 

Logical operators can be used to create more complex conditions to be tested. They include `&`, `|` and `!` (AND, OR and NOT).

### Simple if-else statement

1. Below are two variables. `math_score` stores the score (in percent) of a student on a math test while `passing_score` stores the minimum score needed to pass the test. 
2. In the code box, write an `if-else` statement that prints out `pass` or `fail` according to the math score in relation to the passing score. 
3. Test your statement by changing the value of `math_score` to something less than `passing_score`. Do you observed the expected output?

```{r ifelse_ex, exercise=TRUE, exercise.eval=FALSE}
math_score <- 70
passing_score <- 60

# Write your if-else statement below



```

```{r ifelse_ex-hint}
math_score <- 70
passing_score <- 60

# Write your if-else statement below
if (math_score >= passing_score) {
  print("pass")
} else {
  print("fail")
}

```



### Building more complex conditions

`if-else` statements can also be useful to identify errors in your data. For example, a score should only be between 0 and 100, but a typo at entry could lead to an unacceptable value. 

1. In the box below, write an `if-else` statement that prints out `correct` if the math score value is between 0 and 100, and `incorrect` if the value is outside those bounds. Since we have two bounds to test, 0 and 100, we can use the `&` logical operator to ensure both are tested and true at the same time.
2. Test your code by varying the value of `math_score`.

```{r ifelse_ex2, exercise=TRUE, exercise.eval=FALSE}
math_score <- 705

# Write your if-else statement below



```

```{r ifelse_ex2-hint}
math_score <- 705

# Write your if-else statement below
if (math_score >= 0 & math_score <= 100) {
  print("correct")
} else {
  print("incorrect")
}

```


### Testing for multiple categories

We could also have more than two categories to assign depending on the value of a variable. That is where the `else if` statement comes in. It essentially let's you chain multiple `if` conditional tests, executing the one that is true. To use it, start with a `if` statement, followed by as many `else if` as you need with an optional `else` at the end to catch remaining possibilities. 

1. The box below contains the structure of a test for multiple conditions. The conditions are tested one after the other until one is found true, leading to that section of code being executed. 
2. Complete the structure so that the grade associated with `match_score` is printed following:
* `>=90`: Grade A
* `>=80`: Grade B
* `>=70`: Grade C
* `>=60`: Grade D
* `<60`: Grade F

**Since each tests are done in order, we can end with an `else` statement to cover everything associated with Grade F.**


```{r ifelse_ex3, exercise=TRUE, exercise.eval=FALSE}
math_score <- 85

# Fill the if-else if-else statement below
if (math_score >= 90) {
  
} else if (math_score >= 80) {
  
} else if (math_score >= 70) {
  
} else if (math_score >= 60) {
  
} else {
  
}


```

```{r ifelse_ex3-hint}
math_score <- 85

# Fill the if-else if-else statement below
if (math_score >= 90) {
  print("Grade A")
} else if (math_score >= 80) {
  print("Grade B")
} else if (math_score >= 70) {
  print("Grade C")
} else if (math_score >= 60) {
  print("Grade D")
} else {
  print("Grade F")
}

```


### Nested statements

`if-else` statements can also be applied on top of each other. These are called **nested statements**. You can use these if different range of data values must be categorized or processed differently. Going back to our previous examples, this can be used to first check if the score value is acceptable before either categorizing it or printing an error message. 

Looking at the code in the box below, running it with a `math_score` value of less than 0 or above 100 will inform you the value cannot be graded. With the same code, we can still correctly grade any values between 0 and 100. Give it a try by varying the value of `math_score`.

Note: For code such as this to execute correctly, you need to have matching open and closed parenthesis (like you would in an Excel quation), but you also need matching open and closed `{}` brackets. RStudio will help you by highlighting such errors, but to keep your code clear it is best practice to indent the content of statements by using the `Tab` key at the beginning of a row. 


```{r ifelse_ex4, exercise=TRUE, exercise.eval=FALSE}
math_score <- 850

# Fill the if-else if-else statement below
if (math_score >= 0 & math_score <= 100) {
  if (math_score >= 90) {          # <-- See how code from here...
    print("Grade A")              #
  } else if (math_score >= 80) {  #
    print("Grade B")              #
  } else if (math_score >= 70) {  #
    print("Grade C")              #
  } else if (math_score >= 60) {  #
    print("Grade D")              #
  } else {                        #
    print("Grade F")              #
  }                                # <-- ...to here is indented inwards from the if statement containing it...
} else {
  print("Score outside of 0-100 bounds cannot be graded") # <-- ...same for rows inside each if, else or else if statements.
}


```




## Calling Functions 

Now we'll move on to calling functions in R. A *function* is simply an action that we perform in R. R comes with a number of pre-installed *packages* containing useful functions, but they can also be defined by the user if none exist for the specific task or purpose. Functions are useful in cases where a specific action (say, for example, taking the mean of a vector of numbers) will need to be done repetitively. They can be thought of as analogous to macros in SAS. 

Functions are called by passing an *argument* through them, contained within parentheses: `myFunction()`. Arguments are simply values that the function will be evaluated at (inputs). Thus far, we have used the `print()` functions a handful of times to display the results of our code.

Below, we will practice calling some functions.


### Single argument functions

 1. First, define the vector `x` to contain the numbers 1 through 5. (Hint: there are multiple ways to do this. A shortcut for defining a *sequence* of consecutive numbers in R is to write out the starting and ending values of the sequence, separated by `:`. Alternatively, you may type out each number.)
 
 2. Use the `mean()` function to take the mean of this vector, and store the results in a variable called `mean_x`
 
 3. Print the results.

```{r mean, exercise=TRUE}
x<-c()

```

```{r mean-solution}
x<-c(1:5)
mean_x<-mean(x)
print(x)
```


### Multiple argument functions

Above, we used a function that only required a single argument for evaluation. Below, we will try calling a function that requires multiple arguments.

 1. Use the `seq()` function to define a sequence of numbers from 0 to 50, in increments of 5. The `seq()` function requires three arguments: starting number ("from"), ending number ("to"), and increments ("by"). You can either list the arguments themselves in the specific order above, or list them in whatever order you like by labeling the arguments appropriately (Hint: in this exercise, you would write `from=0` as one of the arguments).
 
 2. Save the output as a variable called `my_seq` and print the results.
 
```{r seq, exercise=TRUE}


```

```{r seq-solution}
my_seq<-seq(0,50,5)

# or 

my_seq<-seq(from=0,to=50,by=5)


print(my_seq)
```


## Importing Packages  

R functions are contained in what are called *packages*: collections of functions that are usually related to one another, or intended for a specific type of analysis. The `demography` package, for example, contains functions useful for demographic analysis (such as the computation of life tables, or forecasting of fertility rates).

R comes with a number of packages already pre-installed, but once you begin to use R more regularly, you'll likely find that you'll need to *import* packages from the [CRAN](https://cran.r-project.org/) (Comprehensive R Archive Network). R allows individuals to create their own packages that, once reviewed, are hosted on an online package repository. To access them, we need to first install them and then call them.

At Statcan, packages are maintained and hosted internally. To learn how to install packages in different environments (AVD, VDI-A, etc.) refer to the following link: https://rpug.pages.cloud.statcan.ca/en/setup/r/r-packages/. If installing a package on your personal computer (or the AVD), you would use the following code:

```{r,eval=FALSE}
install.packages("dplyr")
```

and use the code below to *call* on a package:

```{r,eval=FALSE}
library(dplyr)
```


If a package is not loaded prior to use, then any function within the package that you'd like to use will not be recognized by R.


Lastly, it is possible that two functions with the same name exist in two different packages. For example, the `filter()` function exists within both the `dplyr` and `stats` packages. Which R will choose to recognize depends on the order the package was loaded. If ever the wrong one is defined, you can fix this by re-assignment.

```{r,eval=FALSE}

# Tell R that wou want 'filter' to refer to the filter function in the dplyr package for the current session.

filter<-dplyr::filter

```


Now for some exercises. We will be using the `iris` data set, which comes with R and requires no loading.

 1. Below we would like to use the `select()` function from the `dplyr` package to analyze the `iris` data set. Insert the appropriate line of code below to ensure we can use it (the package itself is already installed).
 
2. Assign the `select()` function to correspond to the `select()` function from the `dplyr` package. 
 
  
```{r package, exercise=TRUE}
iris %>%
  select(-Species)

```

```{r package-solution}
library(dplyr)
iris %>%
  select(-Species)

select<-dplyr::select
```



## Introduction to Data Frames

Data frames are objects in R that contain data: columns are typically variables and rows are observations. Data frames can be created in R directly, or can be imported as .csv, .xlsx, and even SAS files. An alternative name for a data frame within the `tidyverse` is a `tibble`. Tibbles are considered 'tidy' (i.e. optimized) data frame types, and what we will be working with for the remainder of these sessions.

When data frames are imported (see the Learn R section on importing .csv files), it is good practice to inspect them and make sure the data are what you expect. There are a few functions that are particularly useful for this: `head()` will print the first 6 rows of the dataset,  `summary()` presents a statistical summary of the numerical variables (mean, median, etc.) and descriptions of character variables; and `str()` provides a less detailed summary.


1. In the box below, use the `head()` function to print the first 10 rows, rather than 6, of the `iris` data set (hint: the name of the argument used to control the number of rows displayed is `n`).

2. What is the mean of the `Sepal.Length` variable?

```{r df, exercise=TRUE}

```


```{r df-solution}
# First print the first 10 observations
head(iris,n=10)

# Then use the summary() function to view statistical summaries of the variables; the mean is 5.843
summary(iris)
```


### CSVs

In one of the lessons, we learned about CSVs and how to import CSVs as data frames in R. Refer to the appropriate LearnR lesson if you need a refresher. Use your knowledge to answer the questions below.


```{r csv_quiz, echo=FALSE}

quiz(  
  question_checkbox(
    "In a CSV file, values are separated by which symbol?",
    answer("*"),
    answer(",", correct = TRUE),
    answer("space"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  
  question_checkbox(
    "True or false: It is assumed that the first row of a CSV file contains column names",
    answer("TRUE", correct = TRUE),
    answer("FALSE"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  
  question_checkbox(
    "What line of code would we use to import the 'my_data' CSV file using the readr package?",
    answer("read_csv(my_data.csv)"),
    answer("read.csv('my_data.csv')"),
    answer("read_csv('my_data.csv')", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)  


```


### Piping

In R, there are two ways you can perform sequential operations on a number, vector, data set, or any object, really: nested operations, or via *piping*. Piping is specific to the tidyverse, whereas nesting is how you would typically do it in base R. We will demonstrate the difference between the two with a quick example.

Say you would like to take the mean of a vector of numbers from 1 to 5, then take the square root of this, divide this value by 2, and then take the whole thing to the exponent of 4. This is how we would do it in base R:


```{r}
(sqrt((mean(1:5)))/2)^4
```


We work from the inside-out, with each pair of brackets separating one operation from the next. You *can* figure out what's being done, but the code is not the most readable. With the piping operator, given by the symbol `%>%`, we would achieve the same result in the following manner:

```{r}
1:5 %>%
  mean() %>%
  sqrt() %>%
  `/`(2) %>%
  `^`(4)
```


The pipe can be thought of as meaning 'and then': first you define the vector `1:5`, and then you take the mean of this whole thing, and then square root of that, and etc. You do not need to define the arguments within the functions used (for example, `mean()`) *if* the only argument is the initial object itself, because R knows to take the output from the preceding line as input. The result is more lines of code, but easier to read.


1. Perform the following set of operations on the `iris` data set: inspect the first few elements with the `head()` function, select only `Sepal.Length` variable using the `select()` function, then use `summary()` to summarize.


```{r pip, exercise=TRUE}

```


```{r pip-solution}
iris %>%
  head() %>%
  select(Sepal.Length) %>%
  summary()
```



### Column selection

Above, we already got some experience with selecting specific columns within a data frame. More generally, you can list any columns you would like to select with the use of the `select()` function, or, alternative choose to drop specific columns by placing the `-` sign before the column name.


Now for some practice. First, we'll use Use the `colnames()` function to view the columns contained within the`iris` data set

```{r}
colnames(iris)
```


1. From the data set, select only the *Petal.Width* and *Species* columns, using the pipe operator.

2. Achieve the same result as above, except this time by removing the columns we don't need.


```{r sel, exercise=TRUE}

```


```{r sel-solution}
# 1. Select the two columns
iris %>%
  select(Petal.Width, Species)

# 2. Remove the unwanted columns

iris %>%
  select(-Sepal.Length, -Sepal.Width, -Petal.Length)

```
