---
title: "Centre for demography R for beginners"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
## If a package is installed, it will be loaded. If any 
## are not, the missing package(s) will be installed 
## from CRAN and then loaded.

## First specify the packages of interest
packages = c("remotes")
git_packages <- c("learnr")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

## Git packages
gitpackage.check <- lapply(
  git_packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install_github("rstudio/learnr")
      library(x, character.only = TRUE)
    }
  }
)

#require(remotes)
#remotes::install_github("rstudio/learnr")
#require(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


# Codecademy LearnR  - Introduction to R syntax exercises

## R syntax

### Review the basics

In this quiz, we will review the basic elements of R syntax that were covered at the beginning of our session. If you are unsure, 
you can always go review the lesson or use google. Don't stress, you are allowed as many tries as you want. 

```{r basics_quiz, echo=FALSE}
quiz(
  question_numeric(
    "In R, what would be the result of (6a + 42) / (3 * 5) if a = 3?",
    answer(4, correct = TRUE),
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that R follows standard mathematical oder of operations."
  ), 
  
  question_radio(
    "What symbol(s) is used to mark a line as containing comments in R?",
    answer("*"),
    answer("/*"),
    answer("#", correct = TRUE),
    answer("//"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  
  question_checkbox(
    "Select all that would be interpreted as valid strings (combination of characters) when typed in R as shown below. ",
    answer("'hello'", correct = TRUE),
    answer("hello"),
    answer("'déjà vu'", correct = TRUE),
    answer("three dogs"),
    answer("'32'", correct = TRUE),
    answer("3.1415"),
    random_answer_order = TRUE,
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that R needs string sto be clearly identified."
  ),
  
  question_checkbox(
    "Which two of the following are valid R vectors?",
    answer("c(1965, 12, 27)", correct = TRUE),
    answer("('1976', '08' '14')"),
    answer("c(1958, '06', 23)"),
    answer("c('1971', '02', '12')", correct = TRUE),
    answer("c(1986, '8', '6')"),
    random_answer_order = TRUE,
    allow_retry = TRUE,
    incorrect = "Incorrect. Remember that vectors must contain elements of the same type separated by commas AND inside the combine operator c()."
  )
  
)

```


### Vector concepts

For this exercise, you will be writing the code to review vector concepts. In the code box below:

1. Create a `phone_numbers` vector containing three different, 10-digit phone numbers as numeric values. <br>
2. Use `print()` to display the vector content. <br>
3. Using the square brackets `[]` inside the `print()` statement, display only the third element of the vector. <br>
4. **TIP :** you can use vectors `c()` inside square brackets `[]` to access more than one element at a time. Try it for yourself! Using the following syntax will display both the first and third element of your vector: `print(phone_numbers[c(1, 3)])`.

**To see the outputs of your code, click the `>Run Code` button.** 

Note: if you are stuck, click on the hint button.

```{r vectors_ex, exercise=TRUE, exercise.lines = 5}

```

```{r vectors_ex-hint}
# 1
phone_numbers <- c(1234567890, 0987654321, 5647382910)

# 2
print(phone_numbers)

# 3
print(phone_numbers[3])

# 4
print(phone_numbers[c(1, 3)])


```



## if-else statements and operators

Now that we've reviewed some of the basics, let's move on to 

```{r ifelse_ex, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r ifelse_ex-hint}
head(mtcars)
```


## Calling functions

```{r funcall_ex}

```


## Importing packages

```{r packimport_ex}

```
